---
title: "Perform-Case-Control-Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perform-Case-Control-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PhenoGenRLib)
```

# Case Control Analysis by Using Ensembl as A Control

1. We will begin by loading the dataset built-in to this package installation.

```{r load}
nvs <- PhenoGenRLib::linkVariantsWithMetadata(
  metadata = system.file("extdata/huntingtons_datasheet_shortened.csv",
                         package = "PhenoGenRLib"),
  vcfDir = system.file("extdata/", package = "PhenoGenRLib"),
  vcfColName = "vcfs"
)

# Take a look by using View(nvs)!
```

2. This may be useful to use by itself it we were analyzing some specific phenotype we already annotated in the metadata, but for now, we're going to need to extend our dataset by fetching from the Ensembl database. To do this, we'll need rsIDs for as many of our variants as possible. Thus:

```{r Annotate}
mappings <- PhenoGenRLib::mapRsidsForVariants(
  "chromosome",
  variants = huntingtonsVariants,
  offset = 3074680
)
```

Few things to note: We passed in the string `chromosome` to tell the function 
where to find the chromosome identity to match in the Ensembl database. 
Furthermore, we passed in a offset because the VCFs were generated using a 
reference sequence that did not take up the entire chromosome, rather, just a 
specific gene. We therefore need to offset all VCF positions to translate to 
chromosomal positions.

Additionally, it's important to take note of the returned value in that it is 
not a data frame like many of the other functions, rather, a list of two data 
frames. One under the name `$nvs` representing all the variants, similar to 
before, except with the added column of their respective RSIDs. The other is 
under `$rsids` and as the name suggests, is a data frame entirely dedicated to
the rsids giving information on an individual RSID basis.

3. Lastly, let's assume that the VCFs were generated from patients with disease X. We want to know if any of the nucleotide variations are associated statistically with disease X. To do this, we'll compare the frequency of mutations from our hypothetically sick patients, to those that are healthy (Ensembl database).

```{r analysis}
analysis <- PhenoGenRLib::varianceCCAnalysisEnsembl(
  mappings$nvs,
  mappings$rsids,
  totalCaseSamples = 8
)
```
